Extracting The Power of Dependent Types

### 2021-07-13 paper.tex
Small tweaks

It is often desirable to guarantee that a
certain class of errors does not occur in the given program.  These
error classes may include: dereferencing the null pointer, division
by zero, or out of bound indexing.  In most
programming languages, the guarantees that can be checked
statically are very limited.  Nevertheless, moving to a different
language is not always an option.  For example, these languages may
come with great tooling, compilers and/or libraries.  In this paper, we
investigate how to provide strong static guarantees for programs in a
given language, while maintaining the ability to to use existing
compilers and tools.

One very powerful method for enforcing almost any functional property
of a program statically is by using \emph{dependent types}.
In a dependently typed language, properties can be encoded as types,
and if the program typechecks, the property is guaranteed to hold.
If we can extend our existing language with dependent types
(for verification part) and then eliminate them (for using the original
compiler), then our goal is achieved.

While it is possible to add dependent types to an existing language
from scratch, doing so is time-consuming and error-prone.  Instead,
one can also embed the language we want to use into existing
dependently typed languages, which allows us to re-use the typechecker
of the host language to typecheck the embedded language.

@@

Our main objective of the paper is to demonstrate that the
proposed approach works in practice.  Therefore, we guide the reader
through the entire process of: i) shallow embedding of the target
language;  ii) writing programs in this embedding; iii) using
dependent types to statically enforce the properties of interest; 
iv) extract a program in the syntax of the target language
from it.  We run the extracted program using existing
tools and libraries. Essentially, demonstrating that a
shallow embedding in a host language with reflections
suddenly becomes deep.

In order to keep presentation as simple as possible, we have
chosen a very simple target language to embed --- we use a subset
of PostScript~\cite{}, which is a stack-based
language used in the corresponding document format.  While the embedding
may look impractical, the simplicity of the language
allows us to focus on the essence of the proposed approach and avoid unrelated
technical details.  We ensure that many verification patterns found in bigger
languages are covered, implying that the proposed approach scales.
We use Agda as our host language, and the entire paper is Agda script.
This means that the code presented in the snippets has been typechecked
and presented examples ran.

@@

The main advantage of our approach is twofold: first, it allows
you to implement all these things --- as well as the embedded programs
themselves --- \emph{side by side in a single language}, simplifying
the development. And secondly, it allows you to reuse the typechecker
and evaluator of the host language, saving you a lot of work.
While we showcase a single example, the approach has been applied
to several other languages: Kaleidoscope, a minimalist imperative language;
Single assignment C, a high-performance array language;
APL (Array Programming Language), another array language with
heavily overloaded syntax.
For more details on these examples,
refer to~\citet{DBLP:journals/corr/abs-2105-10819}.

>>>>

It is often desirable to guarantee that a
certain class of errors does not occur in the given program.  These
error classes may include: dereferencing the null pointer, division
by zero, or out of bound indexing.  In most
programming languages, the guarantees that can be checked
statically are limited.  Nevertheless, moving to a different
language is not always an option.  For example, these languages may
come with great tooling, compilers and/or libraries.  In this paper, we
investigate how to provide strong static guarantees for programs in a
given language, while maintaining the ability to to use existing
compilers and tools.

One powerful method for enforcing almost any functional property
of a program statically is by using \emph{dependent types}.
In a dependently typed language, properties can be encoded as types,
and if the program typechecks, the property is guaranteed to hold.
If we can extend our existing language with dependent types
(for verification part) and then eliminate them (for using the original
compiler), then our goal is achieved.

While it is possible to add dependent types to an existing language
from scratch, doing so is time-consuming and error-prone.  Instead,
we can \emph{embed} the language we want to use into an existing
dependently typed host language, which allows us to re-use the typechecker
of the host language to typecheck the embedded language.

@@

Our main objective of the paper is to demonstrate that the
proposed approach works in practice.  Therefore, we guide the reader
through the entire process of: i) shallow embedding of the target
language;  ii) writing programs in this embedding; iii) using
dependent types to statically enforce the properties of interest;
iv) extracting a program in the syntax of the target language
from it.  We run the extracted program using existing
tools and libraries. Essentially, we demonstrate that a
shallow embedding in a host language with reflection
suddenly becomes deep.

In order to keep presentation as simple as possible, we have
chosen a very simple target language to embed --- we use a subset
of PostScript~\cite{}, which is a stack-based
language used in the corresponding document format.  While the embedding
may look impractical, the simplicity of the language
allows us to focus on the essence of the proposed approach and avoid unrelated
technical details.  We ensure that many verification patterns found in bigger
languages are covered, showing that the proposed approach scales.

This entire paper is a literate Agda script: all Agda code is typechecked
and all the examples are run while generating the paper.

@@

The main advantage of our approach is twofold: first, it allows
you to implement all these things --- as well as the embedded programs
themselves --- \emph{side by side in a single language}, simplifying
the development. And secondly, it allows you to reuse the typechecker
and evaluator of the host language, saving you a lot of work.
While we showcase a single example, the approach has been applied
to several other languages: Kaleidoscope, a minimalist imperative language;
Single assignment C, a high-performance array language; and
APL (Array Programming Language), another array language with
heavily overloaded syntax.
For more details on these examples,
refer to~\citet{DBLP:journals/corr/abs-2105-10819}.
